---
title: "GEOG 6000 Lab 08b Visualizing spatial Data in R"
author: "Simon Brewer and Blake Vernon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_caption: true
    css: "../style.css"
---

```{r include = FALSE}

knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = "center")

```

The following files will be used in this lab, all available on Canvas:  

- Climate dataset for Western North America: *WNAclimate.csv*
- New York state polygon data in a shapefile: *NY_Data.zip*
- A NetCDF file of global monthly air temperature: *air.mon.ltm.nc*

You will need to download these files from Canvas, and move them from your `Downloads` folder to the `datafiles` folder that you made previously. Make sure to unzip the zip files so that R can access the content. Note that on Windows, you will need to right-click on the file and select 'Extract files'.  


You will also need to make sure the following packages are installed on your computer:  

```{r, eval = FALSE}

pkgs <- c("mapview",
          "raster",
          "RColorBrewer"
          "sf",
          "tmap",
          "viridis")

install.packages(pkgs)

```

```{r}

library(mapview)
library(raster)
library(RColorBrewer)
library(sf)
library(tmap)
library(viridis)

```

# Static Maps with `tmap`

[`tmap`](https://mtennekes.github.io/tmap/) is built on top of `ggplot2` and works in a similar way by building a series of layers with map geometries and elements. We start by using `tm_shape()` to identify the spatial object to be used, and then geometries are added, including filled polygons, borders, legends, etc. 

<br>

## Polygon data

We'll start by making some maps based on the NY shapefile. Start by loading this, then extract only the polygons belonging to Syracuse. 

```{r}

NY8 <- st_read("../datafiles/NY_data/NY8_utm18.shp")

Syracuse <- NY8[NY8$AREANAME == "Syracuse city", ]

```

First, let's make a simple map showing the polygon outlines using `tm_borders()`:

```{r}

tm_shape(Syracuse) + tm_borders()

```

The function `tm_fill()` can be used to define the polygon fill color based on an attribute in the Syracuse data set (`POP8`). Note that this automatically adds a legend within the frame of the figure:

```{r}

tm_shape(Syracuse) + 
  tm_borders() + 
  tm_fill("POP8")

```

The color scale can be changed by setting the `palette` argument in `tm_fill()`. This includes the ColorBrewer scales described above,and the different intervals. For example, to use the 'Greens' palette with percentile breaks:

```{r}

tm_shape(Syracuse) + 
  tm_borders() + 
  tm_fill("POP8", palette = "viridis", style = "quantile")

```

Other map elements can be added. Here we add a longitude/latitude graticule with `tm_graticules()`, a north arrow with `tm_compass()`, and a line of text showing the date the map was made with `tm_credits()`. 

```{r}

tm_shape(Syracuse) + 
  tm_graticules(col = "lightgray") + 
  tm_borders() + 
  tm_fill("POP8", palette = "Greens", style = "quantile") + 
  tm_compass(position = c("left", "bottom")) + 
  tm_credits("2019-10-19", position = c("right", "top"))

```

## Point data

We'll next make some maps with the Western North American site data. As before, we load the data, then convert to an `sf` object. We also load the shapefile of country outlines:

```{r}

wna_climate <- read.csv("../datafiles/WNAclimate.csv")

wna_climate <- st_as_sf(wna_climate, 
                        coords = c("LONDD", "LATDD"),
                        crs = 4326)

countries <- st_read("../datafiles/ne_50m_admin_0_countries/ne_50m_admin_0_countries.shp", 
                     quiet = TRUE)

```

Individual symbols can be plotted on a color scale using `tm_symbols`.

```{r}

tm_shape(wna_climate) + tm_symbols(col = "Jan_Tmp")

```

This takes the same arguments as `tm_fill()` for the color palette. We'll use a red to blue color scale from **RColorBrewer**. The minus sign before the palette name reverses the order of the colors. As there is a large amount of overlap between the sites, we also add an alpha level to make the symbols transparent. 

```{r}

tm_shape(wna_climate) + 
  tm_symbols(col = "Jan_Tmp", alpha = 0.5, palette = "-RdBu") 

```

We'll next add country boundaries from the Natural Earth shapefile loaded earlier. Note that as this is a different spatial object, we have to use `tm_shape()` a second time to reference this, then use `tm_borders()` to add the lines.

```{r}

tm_shape(wna_climate) + 
  tm_symbols(col = "Jan_Tmp", alpha = 0.75, palette = "-RdBu") + 
  tm_shape(countries) + 
  tm_borders(col = "gray")

```

Note that here we've plotted the symbols first to set the plotting window. Unfortunately, this means that the country lines are added on top of the symbols. If we add the country outlines first, then the window is set to the globe. 

```{r eval=FALSE}

tm_shape(countries) + 
  tm_borders(col = "gray") +
  tm_shape(wna_climate) + 
  tm_symbols(col = "Jan_Tmp", alpha = 0.75, palette = "-RdBu")

```

To get around this, we can set the bounding box for the `countries` object using the bounding box of the point data. We'll also move the legend to a better position in the map, and add a title

```{r}

tm_shape(countries, bbox = st_bbox(wna_climate)) + 
  tm_borders(col = "gray") +
  tm_shape(wna_climate) + 
  tm_symbols(col = "Jan_Tmp", alpha = 0.75, palette = "-RdBu") +
  tm_layout(main.title = "W. North America Climate", 
            legend.position = c("left", "bottom"))

```

We can also use one of **tmap**'s preset themes with `tm_style()`:  

```{r}

tm_shape(countries, bbox = st_bbox(wna_climate)) + 
  tm_style("cobalt") + 
  tm_borders(col = "gray") +
  tm_shape(wna_climate) + 
  tm_symbols(col = "Jan_Tmp", alpha = 0.75, palette = "-RdBu") +
  tm_layout(main.title = "W. North America Climate", 
            main.title.size = 1, 
            legend.position = c("left", "bottom")) 

```

Finally, we can write this map to a file using `tmap_save()`. To do this, assign the output of all the **tmap** functions to an R object. Then specify this object and a file name. I've used pdf output here, but you can also use `png`, `tif` or most other common format extensions.

```{r eval=FALSE}

map1 <- tm_shape(countries, bbox = st_bbox(wna_climate)) + 
  tm_borders(col = "gray") +
  tm_shape(wna_climate) + 
  tm_symbols(col = "Jan_Tmp", alpha = 0.75, palette = "-RdBu") +
  tm_layout(main.title = "W. North America Climate", 
            legend.position = c("left", "bottom")) + 
  tm_style("cobalt")

tmap_save(wna_map, "wna_jan_tmp.pdf")

```

## Facetted maps

**tmap** includes a similar facetting function to **ggplot2**. This allows you to make several different versions of the same map split by some grouping factor. A good example of this is where you have repeated measures over time for a set of locations, and you would like to display each time periods separately. You can make all the individual maps, and then add them together with `tmap_arrange()`, or use `tm_facets()`. To illustrate this, we'll load a built in dataset (`metro`) that has population sizes for major metropolitan centers

```{r}
data(metro)
head(metro)
```

This is an `sf` object, with a set of columns with population size for each decade between 1950 and 2030. To plot this as a facetted map, we first need to convert this to long format with one column for area, one for population size and one for year. We'll do this imply by concatenating the different populations sizes, and repeating the location names and year, then put it all into a temporary data frame:

```{r}

pop <- c(metro$pop1950, metro$pop1960, metro$pop1970,
         metro$pop1980, metro$pop1990, metro$pop2000)

name <- rep(metro$name, 6)
year <- rep(seq(1950, 2000, by = 10), each = nrow(metro))

tmp <- data.frame(pop, name, year)

```

Now we'll merge this with the original `sf` object using the location names:

```{r}

metro_long <- merge(metro, tmp, by = "name")

```

The `merge()` function acts in a similar way to SQL's JOIN function. The `name=` argument allows you to specify which columns are to be used to merge. 

Now we can go ahead and plot this. We'll use symbol size to represent the population size and facet by year:

```{r}

tm_shape(metro_long) +
  tm_facets(by = "year") +
  tm_symbols("pop", col = "darkred") +
  tm_shape(countries) +
  tm_borders() +
  tm_layout(main.title = "Metropolitan center population")

```

<br>

## Raster data

In the final example, we'll make figures using the global air temperature dataset. Start by re-reading the data (we also change the name of the layer, to reduce the amount of typing later on...).

```{r}

air_temp <- rotate(raster("../datafiles/air.mon.ltm.nc", varname = "air"))

air_temp

names(air_temp) <- "jan_tmp"

proj4string(air_temp) <- CRS("+init=epsg:4326")

```

Now let's make a plot using `tm_raster()`, which again takes similar options for color palettes. We'll also add the country borders. 

```{r}

tm_shape(air_temp) + 
  tm_raster(col = "jan_tmp", style = "fisher", palette = "-RdBu") +
  tm_shape(countries) + 
  tm_borders() 

```

We can improve this a little by moving the color legend outside of the plotting area. We'll increase the number of color classes to 9, and add a histogram showing the frequency of different values. 

```{r}

tm_shape(air_temp) + 
  tm_raster("jan_tmp", 
            style = "fisher", 
            palette = "-RdBu", 
            legend.hist = TRUE, 
            n = 9) +
  tm_shape(countries) + 
  tm_borders() + 
  tm_layout(legend.outside = TRUE, legend.outside.position = "left")

```

<br>

## Tips and Tricks

For helpful tips and tricks, try:

```{r}

tmap_tip()

```

<br>

# Interactive maps with `mapview`

The [`mapview`](https://r-spatial.github.io/mapview/index.html) package provides functions to produce interactive maps in R by building on the R [`leaflet`](https://rstudio.github.io/leaflet/) package, which provides a low-level interface to the [javascript library](https://leafletjs.com/) of the same name. `mapview` supports multiple spatial data types, including points, lines, polygons, and rasters.

```{r, out.width = '100%'}

mapview(wna_climate, 
        col.regions = "darkred", # fill color
        color = "gray",          # outline color
        alpha.regions = 0.2,     # fill transparency
        alpha = 0.3)             # outline transparency

mapview(Syracuse, 
        col.regions = "darkgreen",
        color = "black")

fn <- system.file("extdata", "kiliNDVI.tif", package = "mapview")

kilimanjaro_NDVI <- raster::stack(fn)[[1]]

mapview(kilimanjaro_NDVI,
        col.regions = viridis::viridis(n = 10)) # supply custom color palette 

```

<br>

Here we will only review a few of the many features that `mapview` provides.

## Basemaps

The default basemap called when using `mapview()` is _CartoDB.positron_. To use other third-party map tiles, we use the `map.types` argument in `mapview()`. For a list of basemap services, see [here](https://leaflet-extras.github.io/leaflet-providers/preview/). For convenience, the `leaflet` package also provides a named vector of these tile services supported by the leaflet plugin. The vector is called `providers` and you can access elements of the vector like you would any other:

```{r}

leaflet::providers[[35]]

leaflet::providers$CartoDB.Positron # for the minimalists out there

```

```{r, out.width = '100%'}

mapview(Syracuse, 
        col.regions = "darkgreen",
        color = "black",
        map.types = c("Stamen.Watercolor", "Stamen.TonerLite"))

```

<br>

## Layers

`mapview` provides for layering much in the same way that `ggplot2` does.  

```{r, out.width = '100%'}

Syracuse_centers <- st_centroid(Syracuse) 

lyr1 <- mapview(Syracuse, 
                col.regions = "darkgreen",
                color = "black")

lyr2 <- mapview(Syracuse_centers,
                col.regions = "darkblue",
                cex = 3) # size of point

lyr1 + lyr2

```

<br>

## Aesthetic Mapping

We can also map attributes of the spatial data to various aesthetics.

```{r, out.width = '100%'}

mapview(Syracuse,
        zcol = "POP8",
        col.regions = viridis::cividis(n = 10),
        alpha.regions = 0.85)

```

```{r, out.width = '100%'}

# breweries in Franconia, Germany
franconia_breweries <- mapview::breweries

# set point size to number of different types of beer served at each brewery
mapview(franconia_breweries, cex = "number.of.types")

```

<br>

## Inset Map

For whatever reason, `mapview` does not currently support interactive inset maps. However, these can still be added using `leaflet` and a little tinkering with the mapview object.

```{r, out.width = '100%'}

add_miniMap <- function(x) leaflet::addMiniMap(x@map) 

imap <- mapview(franconia_breweries, cex = "number.of.types")

add_miniMap(imap)

```

<br>

