---
title: "GEOG 6000 Lab 08b Visualizing spatial Data in R"
author: "Simon Brewer and Blake Vernon"
date: "9/2/2020"
output:
  html_document:
    toc: true
    toc_float: true
    fig_caption: true
    css: "../style.css"
---

```{r include = FALSE}

knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = "center")

```

The following files will be used in this lab, all available on Canvas:  

- Climate dataset for Western North America: *WNAclimate.csv*
- New York state polygon data in a shapefile: *NY_Data.zip*
- A NetCDF file of global monthly air temperature: *air.mon.ltm.nc*

You will need to download these files from Canvas, and move them from your `Downloads` folder to the `datafiles` folder that you made previously. Make sure to unzip the zip files so that R can access the content. Note that on Windows, you will need to right-click on the file and select 'Extract files'.  


You will also need to make sure the following packages are installed on your computer:  

```{r, eval = FALSE}

pkgs <- c("mapview",
          "raster",
          "RColorBrewer"
          "sf",
          "tmap",
          "viridis")

install.packages(pkgs)

```

```{r}

library(mapview)
library(raster)
library(RColorBrewer)
library(sf)
library(tmap)
library(viridis)

```

# Static Maps with `tmap`

[`tmap`](https://mtennekes.github.io/tmap/) is built on top of `ggplot2` and works in a similar way by building a series of layers with map geometries and elements. We start by using `tm_shape()` to identify the spatial object to be used, and then geometries are added, including filled polygons, borders, legends, etc. 

<br>

## Polygon data

We'll start by making some maps based on the NY shapefile. Start by loading this, then extract only the polygons belonging to Syracuse. 

```{r}

NY8 <- st_read("../datafiles/NY_data/NY8_utm18.shp")

Syracuse <- NY8[NY8$AREANAME == "Syracuse city", ]

```

First, let's make a simple map showing the polygon outlines using `tm_borders()`:

```{r}

tm_shape(Syracuse) + tm_borders()

```

The function `tm_fill()` can be used to define the polygon fill color based on an attribute in the Syracuse data set (`POP8`). Note that this automatically adds a legend within the frame of the figure:

```{r}

tm_shape(Syracuse) + 
  tm_borders() + 
  tm_fill("POP8")

```

The color scale can be changed by setting the `palette` argument in `tm_fill()`. This includes the ColorBrewer scales described above,and the different intervals. For example, to use the 'Greens' palette with percentile breaks:

```{r}

tm_shape(Syracuse) + 
  tm_borders() + 
  tm_fill("POP8", palette = "Greens", style = "quantile")

```

Other map elements can be added. Here we add a longitude/latitude graticule with `tm_graticules()`, a north arrow with `tm_compass()`, and a line of text showing the date the map was made with `tm_credits()`. 

```{r}

tm_shape(Syracuse) + 
  tm_graticules(col = "lightgray") + 
  tm_borders() + 
  tm_fill("POP8", palette = "Greens", style = "quantile") + 
  tm_compass(position = c("left", "bottom")) + 
  tm_credits("2019-10-19", position = c("right", "top"))

```

## Point data

We'll next make some maps with the Western North American site data. As before, we load the data, then convert to an `sf` object. We also load the shapefile of country outlines:

```{r}

wna_climate <- read.csv("../datafiles/WNAclimate.csv")

wna_climate <- st_as_sf(wna_climate, 
                        coords = c("LONDD", "LATDD"),
                        crs = 4326)

countries <- st_read("../datafiles/ne_50m_admin_0_countries/ne_50m_admin_0_countries.shp", 
                     quiet = TRUE)

```

Individual symbols can be plotted on a color scale using `tm_symbols`.

```{r}

tm_shape(wna_climate) + tm_symbols(col = "Jan_Tmp")

```

This takes the same arguments as `tm_fill()` for the color palette. We'll use a red to blue color scale from **RColorBrewer**. The minus sign before the palette name reverses the order of the colors. As there is a large amount of overlap between the sites, we also add an alpha level to make the symbols transparent. 

```{r}

tm_shape(wna_climate) + 
  tm_symbols(col = "Jan_Tmp", alpha = 0.5, palette = "-RdBu") 

```

We'll next add country boundaries from the Natural Earth shapefile loaded earlier. Note that as this is a different spatial object, we have to use `tm_shape()` a second time to reference this, then use `tm_borders()` to add the lines.

```{r}

tm_shape(wna_climate) + 
  tm_symbols(col = "Jan_Tmp", alpha = 0.75, palette = "-RdBu") + 
  tm_shape(countries) + 
  tm_borders(col = "gray")

```

We can also use this 

```{r}

tm_shape(wna_climate) + 
  tm_symbols(col = "Jan_Tmp", alpha = 0.75, palette = "-RdBu") + 
  tm_shape(countries) + 
  tm_borders(col = "gray") + 
  tm_style("cobalt")

```

<br>

## Raster data

In the final example, we'll make figures using the global air temperature dataset. Start by re-reading the data (we also change the name of the layer, to reduce the amount of typing later on...).

```{r}

air_temp <- rotate(raster("../datafiles/air.mon.ltm.nc", varname = "air"))

air_temp

names(air_temp) <- "jan_tmp"

proj4string(air_temp) <- CRS("+init=epsg:4326")

```

Now let's make a plot using `tm_raster()`, which again takes similar options for color palettes. We'll also add the country borders. 

```{r}

tm_shape(air_temp) + 
  tm_raster(col = "jan_tmp", style = "fisher", palette = "-RdBu") +
  tm_shape(countries) + 
  tm_borders() 

```

We can improve this a little by moving the color legend outside of the plotting area. We'll increase the number of color classes to 9, and add a histogram showing the frequency of different values. 

```{r}

tm_shape(air_temp) + 
  tm_raster("jan_tmp", 
            style = "fisher", 
            palette = "-RdBu", 
            legend.hist = TRUE, 
            n = 9) +
  tm_shape(countries) + 
  tm_borders() + 
  tm_layout(legend.outside = TRUE, legend.outside.position = "left")

```

<br>

## Tips and Tricks

For helpful tips and tricks, try:

```{r}

tmap_tip()

```

<br>

# Interactive maps with `mapview`

The [`mapview`](https://r-spatial.github.io/mapview/index.html) package provides functions to produce interactive maps in R by building on the R [`leaflet`](https://rstudio.github.io/leaflet/) package, which provides a low-level interface to the [javascript library](https://leafletjs.com/) of the same name. `mapview` supports multiple spatial data types, including points, lines, polygons, and rasters.

```{r, out.width = '100%'}

mapview(wna_climate, 
        col.regions = "darkred", # fill color
        color = "gray",          # outline color
        alpha.regions = 0.2,     # fill transparency
        alpha = 0.3)             # outline transparency

mapview(Syracuse, 
        col.regions = "darkgreen",
        color = "black")

fn <- system.file("extdata", "kiliNDVI.tif", package = "mapview")

kilimanjaro_NDVI <- raster::stack(fn)[[1]]

mapview(kilimanjaro_NDVI,
        col.regions = viridis::viridis(n = 10)) # supply custom color palette 

```

<br>

Here we will only review a few of the many features that `mapview` provides.

## Basemaps

The default basemap called when using `mapview()` is _CartoDB.positron_. To use other third-party map tiles, we use the `map.types` argument in `mapview()`. For a list of basemap services, see [here](https://leaflet-extras.github.io/leaflet-providers/preview/). For convenience, the `leaflet` package also provides a named vector of these tile services supported by the leaflet plugin. The vector is called `providers` and you can access elements of the vector like you would any other:

```{r}

leaflet::providers[[35]]

leaflet::providers$CartoDB.Positron # for the minimalists out there

```

```{r, out.width = '100%'}

mapview(Syracuse, 
        col.regions = "darkgreen",
        color = "black",
        map.types = c("Stamen.Watercolor", "Stamen.TonerLite"))

```

<br>

## Layers

`mapview` provides for layering much in the same way that `ggplot2` does.  

```{r, out.width = '100%'}

Syracuse_centers <- st_centroid(Syracuse) 

lyr1 <- mapview(Syracuse, 
                col.regions = "darkgreen",
                color = "black")

lyr2 <- mapview(Syracuse_centers,
                col.regions = "darkblue",
                cex = 3) # size of point

lyr1 + lyr2

```

<br>

## Aesthetic Mapping

We can also map attributes of the spatial data to various aesthetics.

```{r, out.width = '100%'}

mapview(Syracuse,
        zcol = "POP8",
        col.regions = viridis::cividis(n = 10),
        alpha.regions = 0.85)

```

```{r, out.width = '100%'}

# breweries in Franconia, Germany
franconia_breweries <- mapview::breweries

# set point size to number of different types of beer served at each brewery
mapview(franconia_breweries, cex = "number.of.types")

```

<br>

## Inset Map

For whatever reason, `mapview` does not currently support interactive inset maps. However, these can still be added using `leaflet` and a little tinkering with the mapview object.

```{r, out.width = '100%'}

add_miniMap <- function(x) leaflet::addMiniMap(x@map) 

imap <- mapview(franconia_breweries, cex = "number.of.types")

add_miniMap(imap)

```

<br>





