---
title: "GEOG 6000 Lab 14 Spatial Regression II"
author: "Simon Brewer"
date: "November 12, 2018"
output: 
  pdf_document: 
    number_sections: yes
header-includes:
   - \usepackage{tabularx}
---

```{r echo=FALSE}
options(width=50)
set.seed(1234)
```

In this lab, we will cover some of the more advanced methods for modeling spatial areal data. While you will be working with continuous normal data in the lab, these methods can be used in a generalized form with other data types. You will need several packages with to carry out all the steps listed here: **rgdal**, **spdep**, **spatialreg** and **spgwr**, as well as **classInt** and **RColorBrewer** to make the plots. The lab uses two datasets, both shapefiles, and you will need to download these and unzip them into a directory. 

- Crime rates in Columbus, Ohio (*columbus.zip*)
- Boston housing prices (*boston.tr.zip*)

Things to remember:

- The following font conventions are used:
    - Normal font = text, comments etc
    - `Courier font` = R commands
    - *Italic font* = file names
- R is case-sensitive for variables and functions. Iris is not the same as iris or IRIS.
- `#` indicates comments - no need to type these
- **Remember to change your R working directory to where your files are stored**

**With all the examples given, it is important to not just type in the code, but to try changing the parameters and re-running the functions multiple times to get an idea of their effect.** Help on the parameters can be obtained by typing `help(functionname)` or `?functionname`. 

# Columbus crime dataset
## Reading the dataset
We will again use the crime dataset from Columbus. The dataset is available as a shapefile in *columbus.zip* - you should already have it available, but if not download and extract this before starting.

```{r message=FALSE, results='hide'}
require(rgdal)
col = readOGR("./columbus/columbus.shp")
```

And log-transform the two covariates:

```{r}
col$lINC = log(col$INC)
col$lHOVAL = log(col$HOVAL)
```

## Building the spatial weight matrix
Now build the neighborhood structure and the associated spatial weight matrix. The example below uses the Queen's case definition of neighbors and binary weights, but these can be replaced by other options fairly easily. You will need the **spdep** package for this. 

```{r results='hide', fig.keep='high', results='hide', message=FALSE}
library(spdep)
coords <- coordinates(col)
col.nbq <- poly2nb(col, queen=TRUE)
col.nbq
col.listw = nb2listw(col.nbq)
plot(col, border="white", col="gray")
plot(col.nbq, coords, add=T, col=1, lwd=2)
```

# Spatial filtering

Spatial filtering is modeling approach introduced by Griffith (1978). It uses a eigen-analysis of the spatial weight matrix (the matrix describing which locations are considered to be neighbors) to generate spatial patterns (Moran eigenvectors) that potentially represent the spatial dependence in the data. Using a stepwise approach, it is then possible to see which of these patterns (or set of patterns) represents the autocorrelated error. These can then be used as additional variables in the model to *filter* out the autocorrelation, leaving the coefficients unaffected, and their standard errors robustly estimated. 

For a spatial regression model, these patterns are generated from the product of the spatial weight matrix ($W$) and the matrix of independent variables ($X$), and produces spatial patterns that are tied to the distribution of the variables. The filter can therefore target potential autocorrelation arising from the variables used in the model. 

In R, spatial filtering is carried out as a series of steps:

- Start by building the simple OLS regression model between the dependent and independent variables
- Use the `SpatialFiltering()` function to run the stepwise procedure to select which Moran eigenvectors should be used to filter the data. The selected eigenvector at each step is the one that reduces the Moran's $I$ values for the model the most (i.e. filters the most spatial autocorrelation)
- Rebuild the model including the selected set of Moran eigenvectors

Here, we apply a spatial filter to the Columbus crime dataset. We start by building the OLS model (and run a Moran's $I$ test on the residuals):

```{r results='hide'}
lm.ols <- lm(CRIME ~ lINC + lHOVAL, data=col)
summary(lm.ols)
moran.test(residuals(lm.ols), col.listw)
```

Now build the spatial filter. Note that we have to give the following parameters:

- The neighborhood structure (**not** the spatial weight matrix)
- The weights to be used (`style`)
- The stopping rule (`alpha`). Moran eigenvectors are selected until the $p$-value of the stepwise Moran's $I$ test reach this value
- `ExactEV=TRUE` - calculates exact values for Moran's $I$ expected value and variance. If this is set to FALSE, the values will be calculated using a fast approximation. This can greatly reduce the amount of time required, especially with large datasets.

```{r results='hide', message=FALSE}
library(spatialreg)
sf.err = SpatialFiltering(lm.ols,nb = col.nbq, style = "C",
                          alpha = 0.25, ExactEV = TRUE, data=col)
```

The output from the `SpatialFiltering()` function is an object that contains both information about the stepwise process, and the selected Moran eigenvectors. Summary information about the stepwise process can be obtained by typing:

```{r}
sf.err
```

Each row of the table is one step, and the columns include the following information:

- `Step`: the step number
- `SelEvec`: the eigenvalue of the selected eigenvector
- `MinMi`: Moran's $I$ for residuals with that eigenvector (and previous eigenvectors) included as a filter
- `ZMinMi`: Moran's $I$ for residuals transformed to $z$-score
- `Pr(ZI)`: the $p$-value for the $z$-score
- `R2`: the R$^2$ value of the model with eigenvectors included
- `gamma`: regression coefficient of selected eigenvector in fit

We can plot individual Moran eigenvectors to examine the spatial pattern they represent. The first selected eigenvector is 4:

```{r fig.keep='none'}
library(classInt)
library(RColorBrewer)
col$ME4 = sf.err$dataset[,1]
my.pal = brewer.pal(9, "BuGn")
spplot(col, "ME4", col.regions = my.pal, cuts=8, col="lightgray",
       main = "Moran eigenvector (4)")
```

We now can apply the spatial filter to the original `lm()` model. We first extract the values for the selected eigenvectors, then use this as a term in the model. As there are more than one selected vector, this makes a matrix. This can be included in the call to `lm()` directly, but each vector will be used as an individual covariate in the new regression:

```{r results='hide'}
E.sel <- fitted(sf.err)
lm.sf <- lm(CRIME ~ INC+HOVAL+E.sel, data=col) 
summary(lm.ols)
summary(lm.sf)
```

Note that the coefficients have not changed from the original OLS model, but the significance of them has improved as the autocorrelation has been filtered. In addition, accounting for the extra variance has increased the R$^2$. Finally, we re-run Moran's $I$ test on the model residuals to check that the autocorrelation has been accounted for:

```{r results='hide'}
moran.test(residuals(lm.sf), col.listw)
```

As these are two nested linear models, we can use `anova()` to test whether the additional complexity in the SF model has resulted in a better overall model:

```{r}
anova(lm.ols, lm.sf)
```

And the low $p$-value suggests that we see a significant improvement in the fit. 

# Geographically weighted regression

Geographically weighted regression (GWR) was introduced by Fotheringham and others to deal with issues of autocorrelation and heterogeneity in spatial data. Rather than trying to build a single global model for an entire study region, GWR instead builds many local models within a small window of the total area. The choice of window size is important, as it dictates the number of observations used in each local model, and so the quality of that model. While there has been several papers criticizing this approach as a complete modeling method, it is very useful for exploring potential variation in the relationship between dependent and independent variables, as well as possible misspecification (missing independent variables). We will use it here to explore a dataset of childhood poverty rates in the SE USA. This data is in the file 'south00.shp', in the compressed file 'south.zip'. Download and extract this file, then read it into R:

```{r fig.keep='none', results='hide'}
sus.sp<-readOGR("./south/south00.shp")
## Map of poverty percentage
my.pal <- rev(brewer.pal(9, "RdYlBu"))
spplot(sus.sp, "PPOV", col.regions=my.pal, cuts=8,
       main = "Southern US poverty rates")
```

Now build a neighborhood structure for autocorrelation analysis:
```{r}
south_nb <- poly2nb(sus.sp, queen=T)
south_listw<- nb2listw(south_nb, style="W", zero.policy=TRUE)
```

And run Moran's $I$ to check the degree and significance of autocorrelation in the poverty variable (PPOV).
```{r results='hide'}
moran.test(sus.sp$PPOV, listw=south_listw, zero.policy=T)
```

Next, we build an OLS regression, using the square root of the poverty percentages as dependent variable, and the following variables as independent variables:

- PFHH: Percent female head of household
- PUNEM: Percent unemployed
- PBLK: Percent Black in population
- P65UP: Percent over 65 years old
- METRO: Metropolitan area (binary)
- PHSPLUS: Percent with education beyond high school

Once the model is built, we check for autocorrelation in the residuals with `lm.morantest()`:
```{r results='hide'}
fit1<-lm(SQRTPPOV ~ PFHH + PUNEM + PBLK + P65UP + METRO + PHSPLUS, sus.sp)
summary(fit1)
moran.test(residuals(fit1), south_listw)
```

And we can map the residuals from the model, which show a clear spatial pattern:

```{r eval=FALSE}
## Map residuals
sus.sp$lm.res <- residuals(fit1)
breaks <- classIntervals(sus.sp$lm.res, 9, style = "quantile")
my.pal <- brewer.pal(9, "RdBu") 
spplot(sus.sp, "lm.res", col="transparent", col.regions=my.pal, at=breaks$brks,
       main = "OLS residuals")
```

We now build the GWR model. This requires two steps, the first to assess the best window size, and the second to build and diagnose the local models using this window. The window can be chosen as

- Fixed size: each window will have the same bandwidth or size, so models in data sparse areas will have fewer observations
- Adaptive: rather than setting a single window size, the window for each model is chosen to capture the same number of observations

Here, we will use the second method, by setting the parameter `adapt=TRUE` in the GWR function. This is chosen as there is a lot of variation in the county sizes, so areas with smaller counties will have much a denser set of observations. 

Start by finding the value of `q`: the proportion of points in each window:
```{r eval=FALSE}
library(spgwr)
south.bw<-gwr.sel(SQRTPPOV ~PFHH + PUNEM + PBLK + P65UP + METRO + PHSPLUS,
                  data= sus.sp, adapt = TRUE, gweight=gwr.Gauss,
                  method = "cv", verbose=TRUE) 
south.bw
```

There are several options used here:

- `adapt=TRUE`: The function will use an adaptive bandwidth to ensure approximately the same number of observations are used in each local window to build a model
- `gweight=gwr.Gauss`: Use Gaussian weights to fit each model (closer points will have a greater weight inthe OLS fit)
- `method="cv"`: use cross-validation to test different bandwidth sizes/number of points


We can estimate the approximate number of observations in each window as follows:
```{r eval=FALSE}
print(dim(sus.sp)[1] * south.bw)
```

The outout from this function lists the changing value of `q`, as well as the RMSE results from cross-validation. The value of `q` is changed until no further improvement in the RMSE occurs. Note that we use the output from the previous step to set the window size (`adapt=south.bw`), and we set the window shape to Gaussian (`gweight = gwr.Gauss`)

Now we build the final set of models using the function `gwr()`. 
```{r eval=FALSE}
south.gwr<-gwr(SQRTPPOV ~PFHH + PUNEM + PBLK + P65UP + METRO + PHSPLUS,
               data= sus.sp, adapt = south.bw, gweight = gwr.Gauss, 
               hatmatrix = TRUE)
print(south.gwr)
```

The output of the model gives summary statistics about the range of coefficients across all models built ($n=1387$, the number of counties). In the model diagnostics, you will find an AIC value, which may be used in model comparison, and an R$^2$ value, which should be treated with some caution. 

The output from the `gwr()` function is a large list object with a lot of information in it. One object in this list is a Spatial*DataFrame, with various information about the local models, and we can can now use this to plot out some of the model results. This adds a layer of recursion to the object, so you can access the Spatial object can be referenced by typing (I'd avoid typing this, as it will take a while to print everything held in the object):

```{r eval=FALSE}
south.gwr$SDF
```

And the GWR results are held in the `data` slot:
```{r eval=FALSE}
south.gwr$SDF@data
names(south.gwr$SDF@data)
```

And individual results can be accessed from the `SDF` object:

```{r eval=FALSE}
south.gwr$SDF$localR2
```

First we plot the R$^2$ for each local measure. Areas with low R$^2$ values may indicate model misspecification - missing independent variables (e.g. West Texas). 

```{r eval=FALSE}
breaks <- classIntervals(south.gwr$SDF$localR2, 9, style = "quantile")
my.pal <- brewer.pal(9, "YlOrRd") 
spplot(south.gwr$SDF, "localR2", col="transparent", col.regions=my.pal, at=breaks$brks,
       main = "Local R2 from GWR")
```

And finally we plot the coefficients for one of the independent variables, the percent of the population aged 65 or over (P65UP). 
```{r eval=FALSE}
breaks <- classIntervals(south.gwr$SDF$P65UP, 9, style = "quantile")
my.pal <- brewer.pal(9, "Blues") 
spplot(south.gwr$SDF, "P65UP", col="transparent", col.regions=my.pal, at=breaks$brks,
       main = "Coefficient for P65UP")
```
\newpage

# Spatial hierarchical models

**This is an optional section.**

Hierarchical linear models (HLMs) offer yet another way to account for spatial dependency in your data. R has a couple of libraries that can be used to fit HLMs: **lme4** and **nlme**. We will use the second of these as it has a very flexible scheme for specifying the covariance between errors of a model, including spatial autocorrelation. Install this package if necessary:

```{r eval=FALSE}
install.packages("nlme")
```

And load it:
```{r eval=TRUE}
library(nlme)
```

We'll build a simple hierarchical model with the Columbus data. Our first model will not assume any spatial dependency between locations or model errors, but as this is a hierarchical model, we need to specify a group structure. As there are no natural groups for this, we simply create a dummy group variable, a vector of one's, to place all observations in the same group.

```{r}
col$dummy <- rep(1, dim(col)[1])
```

We can now go ahead and build the model. The function we will use is `lme()`, and the grouping structure of the model is specified using the `|` notation. The specific notation used here `~1 | dummy`, indicates that the intercepts should be treated as a random variable and that `dummy` should be used as a grouping factor:

```{r results='hide'}
col.lme1 = lme(CRIME ~ lINC + lHOVAL, 
              random = ~1 | dummy, data=col@data, method = "ML")
summary(col.lme1)
```

This first model does not add information to the original OLS model, but allows us to now test for and model the spatial autocorrelation. First, we'll do the usual Moran's $I$ test based on the residuals of this model:

```{r results='hide'}
moran.test(residuals(col.lme1), col.listw)
```

Which tells us, not too surprisingly, that the errors show some dependency. The **nlme** package then includes functions to plot the variogram of these errors, so we can see what their structure looks like:

```{r fig.keep='none'}
plot(Variogram(col.lme1, form = ~ Y + X, data = col))
```

The variogram at it's default settings suggest that there is some dependency over fairly short distances. Setting the argument `maxDist` improves the view of this:

```{r}
plot(Variogram(col.lme1, form = ~ Y + X, data = col, maxDist = 10))
```

This suggests that there is spatial covariance, over a distance of about 5 units, and with a small nugget effect (about 0.2). We can now create a new model with `lme()`, where we use the `corr` argument to specify that the errors are expected to spatially covary. The errors can be set using another function `corSpatial`, which will build a geostatistical covariance model. An example of this is given below:

```{r eval=FALSE}
corSpatial(value = c(5, 0.2), form = ~ Y + X, type = 'spherical', nugget = TRUE)
```

This has the following arguments:

- `value`: intial guesses for the range and nugget (if used)
- `form`: the variables defining the coordinates
- `type`: the covariance model (you can see the list of these by typing `help(corSpatial)`)
- `nugget`: whether or not to include a nugget model

Now, we'll add this into the `lme()` function to create our new model (we'll omit the nugget here):
```{r results='hide'}
col.lme2 <- lme(CRIME ~ lINC + lHOVAL, random = ~1 | dummy, 
                data=col@data, method = "ML",
                corr = corSpatial(c(5, 0.2), type="spherical",
                                  form = ~ Y + X, nugget = TRUE))
summary(col.lme2)
```

About halfway through the summary output, you should be able to see the estimated value for the range and nugget. The nugget value is basically zero, so we can re-estimate the model excluding this:

```{r results='hide'}
col.lme2 <- lme(CRIME ~ lINC + lHOVAL, random = ~1 | dummy, 
                data=col@data, method = "ML",
                corr = corSpatial(c(5), type="spherical",
                                  form = ~ Y + X, nugget = FALSE))
summary(col.lme2)
```

Now let's see how this new model compares. First calculate and compare the AIC scores:

```{r}
AIC(col.lme1,col.lme2)
```

Despite the additional complexity of the spatial effect (the new parameter is the estimated range), this provides a decent reduction in AIC. Let's redo the test for residual autocorrelation:

```{r results='hide'}
moran.test(residuals(col.lme2), col.listw)
```

Which surprisingly says that the errors are still highly autocorrelated. The reason for this is that the `residuals()` function here generates the total error term (spatial + random). To get only the random errors, we ask for these to be normalized by the error covariance matrix:

```{r results='hide'}
moran.test(residuals(col.lme2, type="normalized"), col.listw)
```

Which shows that the purely random errors are now no longer autocorrelated. 

\newpage

# Exercise

1. The file *boston.shp* in the compressed file *boston.tr.zip* contains information on house pricing and various social and economic variables for tracts in Boston (the full set of variables is given below. We have explored the local spatial autocorrelation in this data set in a previous lab. Here, you have been asked to produce a statistical model of house prices in Boston, and to explore the various factors that may influence price. You should use the corrected median house values per tract, and log-transform these to remove the skew in the distribution. Unlike the previous questions, there is no single answer to this, instead, you will need to decide on your own modeling strategy among the spatial models we have looked at (including spatial lag and error models, spatial filtering and GWR). At a minimum your answer should include the following:
    + A map of house prices
    + A basic linear regression of house prices against several of the other variables that you consider to be important influences on house price
    + A test of spatial autocorrelation in the residuals of this model
    + A spatial regression model if the autocorrelation is significant, together with the reason that you chose that modeling approach
    + Descriptions of goodness-of-fit of this model
    + A statement describing which variables influence house prices, and the direction, strength and significance of that relationship
    + A brief statement as to whether or not you believe this model is adequate, and what, if anything, might be missing

**Boston housing dataset details:**

  \begin{tabularx}{\linewidth}{| l | X |}
     \hline
		Column header & Variable \\ 
		\hline
		TOWN & Town names \\ 
		TOWNNO & a numeric vector corresponding to TOWN \\ 
		TRACT & Tract ID numbers \\
    LON & Longitude (dd) \\ 
    LAT & Latitude (dd) \\ 
    MEDV & Median values of owner-occupied housing in USD 1000 \\ 
    CMEDV & Corrected median values of owner-occupied housing in USD 1000 \\ 
    CRIM & Per capita crime rate \\ 
    ZN & Proportion of residential land zoned for lots over 25000 sq. ft per town (constant for all Boston tracts) \\ 
    INDUS & Proportion of non-retail business acres per town (constant for all Boston tracts) \\ 
    CHAS & a factor with levels 1 if tract borders Charles River; 0 otherwise \\ 
    NOX & Nitric oxides concentration (parts per 10 million) per town \\ 
    RM & Average numbers of rooms per dwelling \\ 
    AGE & Proportion of owner-occupied units built prior to 1940 \\ 
    DIS & Weighted distance to five Boston employment centres \\ 
    RAD & An index of accessibility to radial highways per town (constant for all Boston tracts) \\ 
    TAX & Full-value property-tax rate per USD 10,000 per town (constant for all Boston tracts) \\ 
    PTRATIO & Pupil-teacher ratios per town (constant for all Boston tracts) \\ 
    B & A numeric vector of $1000*(Bk - 0.63)^2$ where $Bk$ is the proportion of blacks \\ 
    LSTAT & Percentage values of lower status population \\ 
		\hline
\end{tabularx}
Further details can be found here: Harrison, David, and Daniel L. Rubinfeld, Hedonic Housing Prices and the Demand for Clean Air, Journal of Environmental Economics and Management, Volume 5, (1978), 81-102


# R code covered in lab
\begin{tabularx}{\linewidth}{| l | X |}

\hline
R Command & Purpose \\
\hline
\texttt{SpatialFiltering} & Finds the set of Moran eigenvectors that represent spatial autocorrelation in a model\\
\texttt{gwr.sel} & Selects an appropriate window size for GWR. Parameter `adapt=FALSE` selects a global bandwidth (window size). `adapt=TRUE` selects the optimum proportion of observations to include in each window\\
\texttt{gwr} & Builds GWR model for a dataset using window characteristics selected by `gwr.sel()`\\
\hline
\end{tabularx}
